---
title: "Keras Autoencoder and XGBoost Classification in R"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(data.table)
library(keras3)
library(tidymodels)
library(finetune)
library(caret)
tidymodels_prefer()

# Keras 3 supports multiple backends; this project uses TensorFlow explicitly.
use_backend("tensorflow")
seed <- 43111
set_random_seed(seed)
```

## Import data

The analysis uses project-relative paths so it can be reproduced on another machine. See `data/README.md` for the expected files.

```{r}
data_path <- file.path("data", "wave_functions.csv")
if (!file.exists(data_path)) {
  stop("Missing data/wave_functions.csv. See data/README.md for the expected layout.")
}

d <- fread(
  data_path,
  header = FALSE,
  nrows = 800000,
  skip = 1700000
) %>%
  select(1:113) %>%
  rename(Target = V113)

if (ncol(d) != 113) {
  stop("Expected 112 predictors plus one target column.")
}

as_target_factor <- function(x) {
  factor(x, levels = c("1", "0"), labels = c("1", "0"))
}

d <- d %>%
  mutate(Target = as_target_factor(Target)) %>%
  as_tibble()
```

## Sample data

```{r}
set.seed(1821)

df <- d %>%
  slice_sample(prop = 0.003) %>%
  mutate(
    Target = if_else(Target == "0", "Smooth", "Anomaly") %>%
      factor(levels = c("Anomaly", "Smooth"))
  )

dim(df)
```

## Train/test split and preprocessing

The held-out test set is created before any learned preprocessing. The recipe is fitted on training data only and then reused unchanged for test and external-validation data.

```{r}
set.seed(1821)
df_split <- initial_split(df, prop = 0.75, strata = Target)

df_train <- training(df_split)
df_test <- testing(df_split)

basic_rec <- recipe(Target ~ ., data = df_train) %>%
  step_YeoJohnson(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_range(all_numeric_predictors())

basic_prep <- prep(basic_rec, training = df_train)

df_train <- bake(basic_prep, new_data = NULL)
df_test <- bake(basic_prep, new_data = df_test)
```

## Split predictors and targets

```{r}
train_x <- df_train %>% select(-Target)
train_y <- df_train %>% pull(Target)
test_x <- df_test %>% select(-Target)
test_y <- df_test %>% pull(Target)
```

## Keras 3 autoencoder

`set_random_seed()` seeds the R session, Python, NumPy and the selected Keras backend. The autoencoder is trained only on the training partition. Twenty percent of that partition is used internally for early-stopping validation.

```{r}
set_random_seed(seed)

model_encoder <- keras_model_sequential(input_shape = c(ncol(train_x))) |>
  layer_dense(units = 112, activation = "relu") |>
  layer_dense(units = 256, activation = "relu") |>
  layer_dense(units = 6, activation = "relu", name = "embedding") |>
  layer_dense(units = 256, activation = "relu") |>
  layer_dense(units = 112, activation = "relu") |>
  layer_dense(units = ncol(train_x))

model_encoder |> compile(
  loss = "mean_squared_error",
  optimizer = "adam"
)

autoencoder_hist <- model_encoder |> fit(
  x = as.matrix(train_x),
  y = as.matrix(train_x),
  epochs = 100,
  batch_size = 32,
  validation_split = 0.20,
  callbacks = list(
    callback_early_stopping(
      patience = 10,
      restore_best_weights = TRUE
    )
  ),
  verbose = 2
)

embedding_layer <- get_layer(model_encoder, name = "embedding")
embeddings_model <- keras_model(
  inputs = model_encoder$inputs[[1]],
  outputs = embedding_layer$output
)
```

## Build embedding datasets

```{r}
model_data <- predict(embeddings_model, as.matrix(train_x), verbose = 0) %>%
  as.data.frame() %>%
  bind_cols(Target = train_y)

test_set <- predict(embeddings_model, as.matrix(test_x), verbose = 0) %>%
  as.data.frame() %>%
  bind_cols(Target = test_y)

set.seed(1821)
cv_folds <- vfold_cv(
  model_data,
  strata = Target,
  v = 10,
  repeats = 3
)
```

## XGBoost model

```{r}
rcp <- recipe(Target ~ ., data = model_data)

xgb_spec <- boost_tree(
  trees = tune(),
  mtry = tune(),
  min_n = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  loss_reduction = tune(),
  sample_size = tune()
) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

xgb_wf <- workflow() %>%
  add_recipe(rcp) %>%
  add_model(xgb_spec)
```

## Model tuning

Cross-validation tunes the classifier on a fixed representation learned from the training partition. Because the autoencoder is not re-trained inside each XGBoost fold, these resamples are used for **hyperparameter selection only**, not as an unbiased end-to-end performance estimate. The held-out test set remains the primary evaluation set.

```{r}
set.seed(1821)
xgb_rs <- tune_race_anova(
  xgb_wf,
  resamples = cv_folds,
  grid = 6,
  metrics = metric_set(roc_auc),
  control = control_race(verbose_elim = TRUE)
)

plot_race(xgb_rs)
collect_metrics(xgb_rs)
show_best(xgb_rs, "roc_auc")

best_rs <- select_best(xgb_rs, "roc_auc")
best_wf <- finalize_workflow(xgb_wf, best_rs)
```

## Held-out test evaluation

```{r}
model <- fit(best_wf, data = model_data)

test_predictions <- predict(model, test_set)

caret::confusionMatrix(
  test_predictions$.pred_class,
  test_set$Target,
  mode = "everything"
)
```

## Optional external validation

If `data/validation/shock_4.csv` is available, it is transformed with the preprocessing object fitted on the training data and evaluated with the already-trained autoencoder and classifier. Nothing is re-fitted on validation data.

```{r}
validation_path <- file.path("data", "validation", "shock_4.csv")

if (file.exists(validation_path)) {
  shock_4 <- fread(validation_path, header = FALSE)

  if (ncol(shock_4) < 113) {
    stop("Validation data must contain 112 predictors plus one target column.")
  }

  shock_4 <- shock_4[, 1:113]
  names(shock_4)[113] <- "Target"

  shock_4$Target <- factor(
    shock_4$Target,
    levels = c("1", "0"),
    labels = c("Anomaly", "Smooth")
  )

  shock_4 <- as_tibble(shock_4)
  shock_4 <- bake(basic_prep, new_data = shock_4)

  validation_x <- shock_4 %>% select(-Target)
  validation_y <- shock_4 %>% pull(Target)

  validation_set <- predict(
    embeddings_model,
    as.matrix(validation_x),
    verbose = 0
  ) %>%
    as.data.frame() %>%
    bind_cols(Target = validation_y)

  validation_predictions <- predict(model, validation_set)

  caret::confusionMatrix(
    validation_predictions$.pred_class,
    validation_set$Target,
    mode = "everything"
  )
} else {
  message("Optional validation file not found; skipping external validation.")
}
```

## Reproducibility note

`set_random_seed()` controls the main R/Python/NumPy/backend random-number generators. Exact bit-for-bit reproducibility can still depend on hardware, backend version and non-deterministic accelerator operations, so environment versions should also be recorded when publishing benchmark numbers.

## Session information

```{r}
sessionInfo()
```
